Reproducible Research - Peer Assessment 1
========================================================

Load the 'Activity Monitoring' data.
```{r loadData, echo = TRUE}
activity <- read.csv("activity.csv")
```

** What is the mean total number of steps taken each day?**

1. Make a histogram of the total number of steps taken each day.
```{r histogramSteps, echo = TRUE}
totalSteps <- tapply(activity$steps, INDEX = activity$date, FUN = sum)
uniqueDates <- unique(activity$date)
dfSteps <- data.frame(dates = as.factor(uniqueDates), 
                      steps = as.integer(totalSteps))
library(ggplot2)
totalStepsHist <- qplot(dates, data = dfSteps, weight = totalSteps, geom = "histogram") + 
        ylab("Total Number of Steps")
totalStepsHist + theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

2. Calculate and report the __mean__ and __median__ total number of steps taken 
per day.
```{r meanMedianSteps, echo = TRUE}
meanDaySteps <- mean(totalSteps, na.rm = TRUE)
meanDaySteps

medianDaySteps <- median(totalSteps, na.rm = TRUE)
medianDaySteps
```

**What is the average daily activity pattern?**

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis).
```{r timeSeriesPlot, echo = TRUE}
uniqueIntervals <- unique(activity$interval)
dfIntervals <- data.frame(interval = as.integer(), meanSteps = as.numeric())

for (i in uniqueIntervals) {
        intervalSubset <- activity[activity$interval == i,]
        intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
        dfIntervals <- rbind(dfIntervals, c(i, intervalMeanSteps))
}

names(dfIntervals) <- c("interval", "meanNumSteps")
timeSeriesPlot <- ggplot(dfIntervals, aes(interval, meanNumSteps))
timeSeriesPlot + geom_path() + xlab("Interval (min)") + ylab("Mean Number of Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
```{r maxStepsInterval, echo = TRUE}
dfIntervals[which.max(dfIntervals[,2]),]
```
Interval '835' contains the maximum number of steps.

**Imputing missing values**

1. Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs)
```{r numNAs, echo = TRUE}
sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset. 
The strategy does not need to be sophisticated. For example, you could use the 
mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the 
missing data filled in.

```{r replaceNA, echo = TRUE}
activity2 <- activity
activityRows <- as.integer(row.names(activity2))
for (i in activityRows) {
        if (anyNA(activity2[i,]) == TRUE) {
                currentInterval <- activity2[i,3]
                replaceValue <- dfIntervals[dfIntervals$interval == currentInterval, 2]
                activity2[i,1] <- replaceValue
        }
}
```
Each NA value for the 'steps' variable in the 'activity2' dataset was replaced 
by the mean number of steps taken for that respective interval.

4. Make a histogram of the total number of steps taken each day and Calculate 
and report the mean and median total number of steps taken per day. Do these 
values differ from the estimates from the first part of the assignment? What is 
the impact of imputing missing data on the estimates of the total daily number 
of steps?
```{r newHistogram, echo = TRUE}
totalSteps2 <- tapply(activity2$steps, INDEX = activity$date, FUN = sum)
dfSteps2 <- data.frame(dates = as.factor(uniqueDates), steps = as.integer(totalSteps2))
totalStepsHist <- qplot(dates, data = dfSteps2, weight = totalSteps, geom = "histogram") + 
        ylab("Total Number of Steps")
totalStepsHist + theme(axis.ticks = element_blank(), axis.text.x = element_blank())
meanDaySteps2 <- mean(totalSteps2, na.rm = TRUE)
meanDaySteps2
medianDaySteps2 <- median(totalSteps2, na.rm = TRUE)
medianDaySteps2
```

**Are there differences in activity patterns between weekdays and weekends?**

1. Create a new factor variable in the dataset with two levels – “weekday” and 
“weekend” indicating whether a given date is a weekday or weekend day.
```{r weekDayEnd, echo = TRUE}
activity2$date <- as.POSIXlt.date(activity2$date)
activity2$weekday <- weekdays(activity2$date)

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
activity2$dayEnd <- NULL
for (i in activityRows) {
        if (activity2[i,4] %in% weekdays) {
                activity2[i,5] <- "weekday"
        } else {
                activity2[i,5] <- "weekend"
        }
}
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 
5-minute interval (x-axis) and the average number of steps taken, averaged 
across all weekday days or weekend days (y-axis). 
```{r panelPlot, echo = TRUE}

##timeSeriesPlotweekdays <- ggplot(dfIntervals, aes(interval, meanNumSteps))
##timeSeriesPlot + geom_path() + xlab("Interval (min)") + ylab("Mean Number of Steps")
activityWeekdays <- subset(activity2, activity2$V5 == "weekday")
activityWeekends <- subset(activity2, activity2$V5 == "weekend")

weekdayIntervals <- data.frame(interval = as.integer(), meanSteps = as.numeric())
for (i in uniqueIntervals) {
        intervalSubset <- activityWeekdays[activityWeekdays$interval == i,]
        intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
        weekdayIntervals <- rbind(weekdayIntervals, c(i, intervalMeanSteps))
}
names(weekdayIntervals) <- c("interval", "meanNumSteps")

weekendIntervals <- data.frame(interval = as.integer(), meanSteps = as.numeric())
for (i in uniqueIntervals) {
        intervalSubset <- activityWeekends[activityWeekends$interval == i,]
        intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
        weekendIntervals <- rbind(weekendIntervals, c(i, intervalMeanSteps))
}
names(weekendIntervals) <- c("interval", "meanNumSteps")

timeSeriesPlotWeekdays <- ggplot(weekdayIntervals, aes(interval, meanNumSteps)) +
        geom_path() + xlab("Interval (min)") + ylab("Mean Number of Steps") +
        ggtitle("weekday")

timeSeriesPlotWeekends <- ggplot(weekendIntervals, aes(interval, meanNumSteps)) +
        geom_path() + xlab("Interval (min)") + ylab("Mean Number of Steps") +
        ggtitle("weekend")

library(grid)
library(gridExtra)
grid.arrange(timeSeriesPlotWeekdays, timeSeriesPlotWeekends, ncol = 1)


```
