Reproducible Research - Peer Assessment 1
========================================================

### Loading and Preprocessing the data

**Load activity dataset and necessary packages:**
```{r load data, echo = T,results='hide'}
unzip("repdata-data-activity.zip")
activity <- read.csv("activity.csv")
str(activity) ## check out features of the dataset
activity$date <- as.Date.factor(activity$date) ## convert dates to class Date

library(ggplot2)
library(plyr) 
```

### What is mean total number of steps taken per day?

**1. Make a histogram of the total number of steps taken each day.**
```
        Here I created a separate dataset for the number of steps taken each
        day. From this dataset the histogram was created.
```
```{r histogram totalSteps, echo = T, warning=F}
uniqueDates <- unique(activity$date)
totalSteps <- tapply(activity$steps, INDEX = activity$date, FUN = sum)
dfSteps <- data.frame(uniqueDates, totalSteps)
qplot(totalSteps, data=dfSteps, geom="histogram",
      xlab = "Total Number of Days",
      ylab = "Number of Days (Non Consecutive)",
      main = "Histogram - Total Number of Steps Per Day")
```

**2. Calculate and report the mean and median total number of steps taken**
**per day.**
```
        I take this instruction as: report the mean and median number of steps 
        taken over the period of the two months.
```
```{r calc. Mean and Median steps, echo = T, eval=F}
mean(dfSteps$totalSteps, na.rm = T)
median(dfSteps$totalSteps, na.rm = T)
```
```
        Mean number of total steps taken each day: 10766.19
        Median number of total steps taken each day: 10765
```

### What is the average daily activity pattern?

**3. Make a time series plot (i.e. type = "l") of the 5-minute interval**
**(x-axis) and the average number of steps taken, averaged across all** 
**days (y-axis).**

```{r time series plot, echo = T}
uniqueIntervals <- unique(activity$interval)
dfIntervals <- data.frame(interval = as.integer(), meanSteps = as.numeric())

for (i in uniqueIntervals) {
    intervalSubset <- activity[activity$interval == i, ]
    intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
    dfIntervals <- rbind(dfIntervals, c(i, intervalMeanSteps))
}

names(dfIntervals) <- c("interval", "meanNumSteps")
timeSeriesPlot <- ggplot(dfIntervals, aes(interval, meanNumSteps))
timeSeriesPlot + geom_path() + xlab("Interval (min)") + 
        ylab("Mean Number of Steps") + 
        ggtitle("Mean Number of Steps per Interval")
```

**4. Which 5-minute interval, on average across all the days in the dataset** 
**contains the maximum number of steps?**

```{r max Steps, echo = T, eval=F}
dfIntervals[which.max(dfIntervals[,2]),] #This outputs a table with max steps
```
```
        The interval with the maximum average number of steps is interval 835.
```
### Imputing missing values

5. Calculate and report the total number of missing values in the dataset 
(i.e. the total number of rows with NAs)

```{r calculate NAs, echo=T, eval=F}
sum(is.na(activity$steps))
```

```
        There are 2304 missing values in the activity dataset.
```

**6. Devise a strategy for filling in all of the missing values in the** 
**dataset.The strategy does not need to be sophisticated. For example, you** 
**could use the mean/median for that day, or the mean for that 5-minute** 
**interval, etc.**
```
        For each missing value, the mean number of steps for that particular
        interval will be used (i.e. if there is an NA value during a "5" 
        interval, for which the mean number of steps is 0.3396226, it will be 
        replaced with 0.3396226.)
```        

**7. Create a new dataset that is equal to the original dataset but with the** 
**missing data filled in.**


```{r imputed dataset, echo = T}
activityNew <- activity
activityRows <- as.integer(row.names(activityNew))
for (i in activityRows) {
        if (anyNA(activityNew[i, ]) == TRUE) {
                currentInterval <- activityNew[i, 3]
                replaceValue <- dfIntervals[dfIntervals$interval == currentInterval, 
                2]
                activityNew[i, 1] <- replaceValue
        }
}
str(activityNew)
sum(is.na(activityNew))
```

**8. Make a histogram of the total number of steps taken each day and** 
**Calculate and report the mean and median total number of steps taken per** 
**day.Do these values differ from the estimates from the first part of the** 
**assignment? What is the impact of imputing missing data on the estimates of** 
**the total daily number of steps?**

```{r newHistogram totalSteps, echo=T}
uniqueDates2 <- unique(activityNew$date)
totalSteps2 <- tapply(activityNew$steps, INDEX = activityNew$date, FUN = sum)
dfSteps2 <- data.frame(uniqueDates2, totalSteps2)
qplot(totalSteps2, data=dfSteps2, geom="histogram",
      xlab = "Total Number of Days",
      ylab = "Number of Days (Non Consecutive)",
      main = "Histogram - Total Number of Steps Per Day (with Imputed Values)")

```

```
        The Histogram appears unchanged in its distribution of values. Adding
        the imputed values, from the method chosen above, may amplify the 
        distribution already given by the original data points, but does not
        appear to noticeably change it.
```

```{r new Mean and Median, echo = T, warning=FALSE}
mean(dfSteps2$totalSteps, na.rm = T)
median(dfSteps2$totalSteps, na.rm = T)
```
```
        The mean and median of the total number of steps, with the new imputed 
        values in the 'activityNew' dataset, change very little. By 
        substituting NA value with the mean of each respective interval, it
        would make sense that the mean and median changed very little.
```

### Are there differences in activity patterns between weekdays and weekends?

**9. Create a new factor variable in the dataset with two levels – “weekday”**
**and “weekend” indicating whether a given date is a weekday or weekend day.**
```
        It is not specified whether to use the orignal dataset, or the new dataset
        with imputed values, so for this part, I will be using the 'activityNew' 
        dataset.
```
```{r weekday variable, echo = T}
activityNew$weekday <- weekdays(activityNew$date)
days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

activityNew$dayEnd <- rep(NA, length(activityNew$weekday))
for (i in activityRows) {
        if (activityNew[i, 4] %in% days) {
                activityNew[i, 5] <- "weekday"
        } else {
                activityNew[i, 5] <- "weekend"
        }
}
```

**10. Make a panel plot containing a time series plot (i.e. type = "l") of the** 
**5-minute interval (x-axis) and the average number of steps taken, averaged** 
**across all weekday days or weekend days (y-axis).**

```{r timeSeriesPlot, echo=TRUE}
weekdaySet <- subset(activityNew, activityNew$dayEnd == "weekday")
weekendSet <- subset(activityNew, activityNew$dayEnd == "weekend")

weekdayIntervals <- data.frame(interval = as.integer(),
                               meanSteps = as.numeric())

for (i in uniqueIntervals) {
    intervalSubset <- weekdaySet[weekdaySet$interval == i, ]
    intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
    weekdayIntervals <- rbind(weekdayIntervals, c(i, intervalMeanSteps))
}
names(weekdayIntervals) <- c("interval", "meanNumSteps")

weekendIntervals <- data.frame(interval = as.integer(),
                               meanSteps = as.numeric())

for (i in uniqueIntervals) {
    intervalSubset <- weekendSet[weekendSet$interval == i, ]
    intervalMeanSteps <- mean(intervalSubset$steps, na.rm = TRUE)
    weekendIntervals <- rbind(weekendIntervals, c(i, intervalMeanSteps))
}
names(weekendIntervals) <- c("interval", "meanNumSteps")

timeSeriesPlotWeekdays <- ggplot(weekdayIntervals,
                                 aes(interval,meanNumSteps)) + 
        geom_path() +
        ylab("Mean Number of Steps") + 
        ggtitle("weekday")

timeSeriesPlotWeekends <- ggplot(weekendIntervals, 
                                 aes(interval, meanNumSteps)) + 
        geom_path() +
        xlab("Interval (min)") + 
        ylab("Mean Number of Steps") + 
        ggtitle("weekend")

library(grid)
library(gridExtra)
grid.arrange(timeSeriesPlotWeekdays, timeSeriesPlotWeekends, ncol = 1)
```


```
        This panel plot shows the mean number of steps taken for each interval
        on weekdays and weekend days. It appears that, on average, there are 
        more steps being taken during the middle of the day on weekend days.
```
